Aggressive Scheduling for Numerical Programs

### 2012-05-01 report/draft.tex
Related work.

\begin{verbatim}
+( ) OSKI
+( ) FFTW
+( ) ATLAS/PhiPack
+( ) [Yotov] discusses how ATLAS only needed fine tuning and many of the course
  adjustments could be determined analytically.
+( ) Lam's software pipelining technique
\end{verbatim}

Lam's software pipelining technique~\cite{Lam:04}.

>>>>

Much research has been done to automatically generate the best implementation
for a given architecture using tunable parameters. These parameters are
related to standard techniques such as loop transformations, loop tilling and loop unrolling.
In these research systems, there is a procedure that searches for the best combination of optimizations
that will generate the most performant implementation.

A very well-known system is ATLAS \cite{Whaley00automatedempirical} (Automatically Tuned Linear Algebra Software).
It automatically generates high performant numerical libraries known as the Basic Linear Algebra Subprograms (BLAS)~\cite{Lawson:79}
using \emph{global search} over the optimization parameters. Many programs are generated with different combinations
and then run on the actual hardware to determine the best values that give the best performance.
ATLAS can also make use of hand-coded BLAS routines in the case that the best solution performs worse than the hand-coded version.

FFTW3~\cite{Frigo05thedesign} is also similar to ATLAS, although FFTW3 targets the implementation of the Discrete Fourier Transform.
FFTW3 uses the concept of a \emph{planner} that receives as input the shape of the input data (array sizes, memory layouts, etc).
The planner returns a \emph{plan} that is able to solve the problem using data that conforms to the initial shape. The planner measures
the execution time of different plans and selects the best one. FFTW3 results show that the best plan performs very close
to what an expert would do.

While both ATLAS and FFTW3 use the same global search principle, Yotov~\cite{Yotov05issearch} showed that it is possible to get
similar results just by using traditional \emph{analytical models}. In his work, he modified the search engine of ATLAS to use
a model parameter estimator that receives hardware information (L1 cache size, number of registers, etc) and outputs
one implementation of the target algorithm that should be best suited for the input parameters. Yotov wanted to show that
traditional compilers should be able to get similar results as ATLAS without the need to execute multiple versions of the same
algorithm to find the best solution. He showed that model driven optimization can achieve very similar results, without the need
to do extensive and time consuming empirical search.

\begin{comment}
+( ) OSKI
+( ) FFTW
+( ) ATLAS/PhiPack
+( ) [Yotov] discusses how ATLAS only needed fine tuning and many of the course
  adjustments could be determined analytically.
+( ) Lam's software pipelining technique
\end{comment}

Software pipelining was a technique invented by Patel and Davidson for scheduling hardware pipelines. Later Rau and Glaeser~\cite{Rau:81}
were able to implement software pipelining in a compiler. In her seminal paper, Lam~\cite{Lam:88} showed that it is possible to implement
software pipelining without hardware support and just instruction-level parallelism. This can be accomplished by using modulo variable
expansion and unrolling of the loop code a small number of times, a technique called \emph{modulo renaming}.
Lam also noted that inter-iteration dependences in a loop can create cycles in the dependence graph and therefore constraint the
scheduling of all other instructions. Lam presented some techniques that deal with this problem.
